---
title: "Example: Data Frame Basics"
author: "Thomas Nauss"
date: "19 Oktober 2017"
output: 
  html_document: 
    keep_md: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path='{{ site.baseurl }}/assets/images/rmd_images/e01-01/')
```

Data frames are one of the most heavily used data structures in R.

## Creation of a data frame
A data frame is created from scratch by supplying vectors to the the `data.frame`
function. Here are some examples:
```{r}
x <- c(2.5, 3.5, 3.4)
y <- c(5, 10, 1)
my_df <- data.frame(x, y)
my_df
colnames(my_df) <- c("Floats", "Integers")

my_other_df <- data.frame(X = c(2, 3, 4), Y = c("A", "B", "C"))
my_other_df
```
The `colnames` function allows to supply column names to an existing data
frame. Alternatively, the column names can be set within the `data.frame` 
function by assign the vector elements to a variable (X and Y in the
second example above).


## Dimensions of a data frame
To get the dimensions of a data frame, use the `ncol` (number of columns), 
`nrow` (number of rows) or `str` (structure) function:
```{r}
ncol(my_other_df)
nrow(my_other_df)
str(my_other_df)
```
## Displaying and accessing the content of a data frame
The content of a data frame is accessed by either a position information 
given in square brackets (e.g. `df[3,4]`) or a column name given after a $ sign
(e.g. df$columnName). Here's an example:



```{r}
my_other_df[1,]  # Shows first row
my_other_df[,2]  # Shows second column
my_other_df$Y  # Shows second column
```

If position information is used, the ordering matters. If you think of a data
frame like a table, then the following applies:

 * In a 1-D data frame, the first dimension is the row
 * In a 2-D data frame, the first dimension is the row, the second the column

Higher dimensions follow the same logic.

Here are some possible combinations:

 * Single row, all columns: `df[x,]` with $x \in \text{number of rows}$
 * Single column, all rows: `df[,y]` with $x \in \text{number of columns}$
 * Single row and column: `df[x,y]` with $x, y \in \text{number of rows, columns}$
 * All except one row, all columns: `df[-x,y]` with $x \in \text{number of rows}$
 * Selected rows, all columns: `df[c(x1, x2, x3),]` with $x1, x2, x3 \in \text{number of rows}$
 * Continous rows, all columns: `df[c(x1:x2),]` with $x1, x2 \in \text{number of rows}$

In summary, dimensions like rows or columns that should be selected have positive 
numbers, such that should be hidden have negative numbers and if all entries of
a dimension should be selected, one just leaves the field empty. If more than
one dimension should be shown/hidden, one has to supply this information by a 
vector which is defined by the `c` function.

```{r}
my_other_df[c(1,3),]  # Shows rows 1 and 3
my_other_df[c(1,2),]  # Shows rows 1 to 2
```

If you are interested in the first or last rows, you can also use the 'head' or
'tail' command. The default number of lines is 5 but this can be changed by the
second argument. Let's have a look at the first two rows:
```{r}
head(my_other_df, 2)
```

And now on the last two rows:
```{r}
tail(my_other_df, 2)
```

## Changing, adding or deleting an element of a data frame
In order to change an element of a data frame (individual value or entire
vectors like rows or columns), you have to access it following the logic above.
To add or delete a column, you have to supply/remove a vector to the specified
position.

Other (more individual) changes or adding rows will be covered later. 
```{r}
# overwrite an element
my_other_df$X[3] <- 400  # same as my_other_df[3,1] <- 400
my_other_df

# change an entire dimension
my_other_df[,1] <- c("200", "300", "401")  # same as my_other_df$X <- 400
my_other_df

# add a new column
my_other_df$z <- c(255, 300, 100)
my_other_df

# delete a column
my_other_df$z <- NULL
my_other_df

```
As for lists, to actually delete an element, it has to be set to `NULL`.

For more information have a look at e.g. the respective [data type](http://www.statmethods.net/input/datatypes.html) site at Quick R. There you will also find an overview on how to get [information about an object](http://www.statmethods.net/input/contents.html). Of course, looking into the package documentation or search the web is always a good idea, too.