---
title: "Example: If-then-else"
author: "Thomas Nauss"
date: "19 Oktober 2017"
output: 
  html_document: 
    keep_md: yes
---

If-then-else statements are the controlling structures in each program. The most
simplest form is:
```{r}
a <- 5.0
b <- 10.0
if (a < b) {
  print("a is smaller than b")
}
```
Such an if-statement can be extended by n if-else statements and - optionally - 
an else-statement. To illustrate this, we start with another if-statement as the
one above but include it later:
```{r}
if (a < b/2) {
  print("a is much smaller than b")
} else if (a < b) {
  print("a is smaller than b")
} else if (a == b) {
  print("a equals b")
} 
```
In such a structure, the first if-statment is evaluated. If it is true, all 
following else-if- (or else-)statements are skiped and the execution continues
below the decission structure.

The above example would not print anything if a is larger than b. For this case,
other else-if-statements could be used but if no differentiation should be made
for any case of a larger than b, this can also be solved by a final 
else-statement.
```{r}
a <- 15.0
b <- 10.0
if (a < b/2) {
  print("a is much smaller than b")
} else if (a < b) {
  print("a is smaller than b")
} else if (a == b) {
  print("a equals b")
} else {
  print("a is larger than b.")
}
```
Of course, boolean expressions could also be used within such structures:
```{r}
c <- 20.0
if (a < b & a == c) {
  print("a is smaller than b and equal to c")
} else if (a > b & a == c) {
  print("a is larger than b and equal to c")
} else {
  print("a non-equal to c.")
}
```
The above examples document if-statements which are executed if the condition is
true since this is the default action for something like `a < b`. Of course,
one can also check if a condition is false:
```{r}
test <- a < b
if (test == FALSE) {
  print("a is larger than b")
} else {
  print("a is smaller than or equal b.")
}
```