---
title: "Example: Aggregation Statistics"
author: "Thomas Nauss"
date: "19 Oktober 2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path='{{ site.baseurl }}/assets/images/rmd_images/e01-01/')
source("C:/Users/tnauss/permanent/edu/mpg-data-analysis/moer-mpg-data-analysis/staging/examples/000_set_environment.R")
```

## Summarizing a data set

The most straight forward function which returns some aggregated statistical
information about a data set is `summary`.
```{r}
a <- c("A", "B", "C", "A", "B", "A", "A")
b <- c("X", "X", "X", "X", "Y", "Y", "Y")
c <- c(1, 2, 3, 4, 5, 6, 7)
d <- c(10, 20, 30, 40, 50, 60, 70)
df <- data.frame(Cat1 = a, Cat2 = b, Val1 = c, Val2 = d)
summary(df)
```
It provides the minimum, maximum, mean and median value of each numerical column
along with the first and third quartile (i.e. the values separating 25% from 75%
and 75% from 100% of the data). Non-numeric columns are summarized by counting
the individual values (which will be ommitted after the first few rows so it
is only informative for columns with just a few different character values).

The information of `summary` can also be obtained by specific functions but these
will return only the same numbers if applied to a single column. For example the
`min` function will return the minimum value of all values supplied to it. This
leads to a return of 1 no matter if it is applied to column Val1 or to columns
Val1 and Val2:
```{r}
min(df[,3])
min(df[,3:4])
```
The other functions included in summary are
```{r}
max(df$Val1)
quantile(df$Val1,  probs = c(0.25, 0.75))
mean(df$Val1)
median(df$Val1)
```


## Aggregated statistics
If you are interested in aggregated statistics in terms of sums or means across
columns or rows, you can use one of the `rowSums`, `rowMeans`, `colSums`, or 
`colMeans` functions.
```{r}
df  # Just print the data frame so you can easily understand the results
rowSums(df[,3:4])
rowMeans(df[,3:4])
colSums(df[,3:4])
colMeans(df[,3:4])
```

For other summary statistics across rows or if you want to define a grouping
variable which controls the summary statistics, the `aggregate` function is very
helpfull.
```{r}
aggregate(df[,3:4], by = list(df$Cat1), FUN = median)
```
Please note that the grouping elements (parameter "by") have to supplied as list
so if you want to use a column of a data frame, use the `list` function to convert
the factor to a list.

Of course, one can also use more than one grouping variable.
```{r}
aggregate(df[,3:4], by = list(df$Cat1, df$Cat2), FUN = sum)
```

In some cases it is usefull to construct a specific aggregation key, so think of
that easy possibility if you run into problems. As one example: imagine that the
aggregate function can not use more than one grouping variable. To get the same
result as in the line above, one can define a key consisting of the entries of
column Cat1 and Cat2 (sorting is different but who cares):
```{r}
df$key <- paste0(df$Cat1, df$Cat2)
aggregate(df[,3:4], by = list(df$key), FUN = sum)
```
