---
title: "Example: Data Frame Basics"
author: "Thomas Nauss"
date: "19 Oktober 2017"
output: 
  html_document: 
    keep_md: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path='{{ site.baseurl }}/assets/images/rmd_images/e01-01/')
```

Data frames are one of the most heavily used data structures in R.

## Creation of a data frame
A data frame is created from scratch by supplying vectors to the `data.frame`
function. Here are some examples:
```{r}
x <- c(2.5, 3.5, 3.4)
y <- c(5, 10, 1)
my_df <- data.frame(x, y)
my_df
colnames(my_df) <- c("Floats", "Integers")

my_other_df <- data.frame(X = c(2, 3, 4), Y = c("A", "B", "C"))
my_other_df
```
The `colnames` function allows you to supply column names to an existing data frame. You can also set column names directly in `data.frame()`, as shown by the named arguments `X` and `Y` above.


## Dimensions of a data frame
To get the dimensions of a data frame, use the `ncol` (number of columns), 
`nrow` (number of rows) or `str` (structure) function:
```{r}
ncol(my_other_df)
nrow(my_other_df)
str(my_other_df)
```
## Displaying and accessing the content of a data frame
Access values in a data frame using indices in square brackets (e.g., `df[3,4]`) or a column name after a `$` sign (e.g., `df$columnName`). Here is an example:


```{r}
my_other_df[1,]  # Shows first row
my_other_df[,2]  # Shows second column
my_other_df$Y  # Shows second column
```

A data frame has rows and columns. With two indices, use `df[rows, columns]`: the first index selects rows, and the second selects columns. Leaving an index empty selects all entries in that dimension. With a single index, `df[index]` selects columns.

In the examples below, `df` represents a data frame, `x` represents a row number, and `y` represents a column number.

Here are some possible combinations:

 * Single row, all columns: `df[x,]`
 * Single column, all rows: `df[,y]`
 * Single row and column: `df[x,y]`
 * All except one row, all columns: `df[-x,]`
 * Selected rows, all columns: `df[c(x1, x2, x3),]`
 * Consecutive rows, all columns: `df[c(x1:x2),]`

Use positive indices to select rows or columns and negative indices to exclude them. To select or exclude several rows or columns, supply a vector of indices using `c()`.

```{r}
my_other_df[c(1,3),]  # Shows rows 1 and 3
my_other_df[c(1,2),]  # Shows rows 1 to 2
```

To display the first or last rows, use `head` or `tail`. By default, these functions display six rows. You can change this number with the second argument. Let us look at the first two rows:

```{r}
head(my_other_df, 2)
```

And now let's look at the last two rows:
```{r}
tail(my_other_df, 2)
```

## Changing, adding or deleting an element of a data frame
In order to change an element of a data frame (individual value or entire
vectors like rows or columns), you have to access it following the logic above.
To add or delete a column, you have to supply/remove a vector to the specified
position.

Other more specific changes will be covered later.
```{r}
# overwrite an element
my_other_df$X[3] <- 400  # same as my_other_df[3,1] <- 400
my_other_df

# change an entire column
my_other_df[,1] <- c("200", "300", "401")  # same as my_other_df$X <- c("200", "300", "401")
my_other_df
```

The values are in quotation marks, so `X` becomes a character column.

```{r}
# add a new column
my_other_df$z <- c(255, 300, 100)
my_other_df

# delete a column
my_other_df$z <- NULL
my_other_df

```
As for lists, to actually delete an element, it has to be set to `NULL`.

For more information, see the accompanying Base R course on [data types](https://geomoer.github.io/moer-base-r/unit02/unit02-01_Intro.html) and [object types](https://geomoer.github.io/moer-base-r/unit03/unit03-01_Intro.html). Package documentation is another useful resource.
