---
title: "Example: Merging"
author: "Thomas Nauss"
date: "19 Oktober 2017"
output: 
  html_document: 
    keep_md: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path='{{ site.baseurl }}/assets/images/rmd_images/e01-01/')
```

When thinking about combining two data frames one has to distinguish between
merging them by the values given in a specific column or consecutively putting
them together just rows after rows.

For the following examples, we will use these two data frames.
```{r}
x <- c(1, 2, 3, 4, 5)
y <- c(1.4, 2.5, 3.6, 4.0, 5.5)
z <- c("A", "B", "C", "D", "E")
df1 <- data.frame(X = x, Y = y, Z = z)
df1

y <- c(14, 25, 36, 40, 55)
x <- c(100, 200, 300, 400, 500)
z <- c("A", "C", "E", "D", "B")
df2 <- data.frame(X = x, Y = y, Z = z)
df2
```

To combine two data frames, use the `rbind` function provided that both
data frames have the same column names and the same number of columns.
```{r}
dfr <- rbind(df1, df2)
dfr
```
Note, that it is not necessary that the columns are in the same order since the
concatenation is done by the column names, not the column sequence:
```{r}
df3 <- df2[,c(1,3,2)]
df3
dfr <- rbind(df1, df3)
dfr
```


To merge two data frames, you have to specify the column containing the values
which should be used in the matching. In this example, we use the values of
the column Z to merge the two data frames.
```{r}
dfc <- merge(df1, df2, by = "Z")
dfc
```
Since both data frames have identical column names, X and Y is added to the 
column names in the resulting data frame to indicate if the column is from the
first (i.e. X) or second (i.e. Y) data frame. You can easily rename the columns
using the `colnames` function.

If the columns which should be used for mergin have different names, no problem:
just supply the column names separately for the first (i.e. X) and second (i.e. Y)
data frame:
```{r}
colnames(df2) <- c("H", "I", "J")
dfc <- merge(df1, df2, by.x = "Z", by.y = "J")
dfc
```
Since not only the names of the columns used for merging but all column names are
different, no X or Y is added in the output column names.